---
title: "STAT 209: Project Presentation"
author: "Mahnaz Pirgazi and Behzad FallahiFard"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-depth: 2
    number-sections: false
    smooth-scroll: true
    fig-align: center
    df-print: paged
    code-copy: true
    code-overflow: wrap
execute:
  warning: false
  message: false
---

<style>
:root {
  --main-blue: #1f4e79;
  --light-blue: #eef5fb;
  --soft-gray: #f7f9fb;
  --border-gray: #d9e2ec;
}

body {
  font-size: 17px;
  line-height: 1.55;
}

.quarto-title-block .title {
  color: var(--main-blue);
  font-weight: 750;
  letter-spacing: 0.2px;
}

h1 {
  color: var(--main-blue);
  font-weight: 750;
  border-bottom: 3px solid var(--border-gray);
  padding-bottom: 0.35rem;
  margin-top: 1.4rem;
}

h2, h3 {
  color: var(--main-blue);
  font-weight: 700;
  margin-top: 1.2rem;
}

p, li {
  font-size: 1.02rem;
}

ul, ol {
  margin-bottom: 1rem;
}

code {
  color: #8a3a00;
  background-color: #fff4e6;
  border-radius: 4px;
  padding: 0.08rem 0.22rem;
}

pre code {
  color: inherit;
  background-color: transparent;
  padding: 0;
}

.cell-output-display, .figure {
  margin-top: 0.7rem;
  margin-bottom: 1.2rem;
}

.table, table {
  font-size: 0.95rem;
}

.dataTables_wrapper {
  font-size: 0.92rem;
}

.dataTables_wrapper .dataTables_caption,
caption {
  caption-side: top;
  text-align: left;
  color: var(--main-blue);
  font-weight: 700;
  margin-bottom: 0.35rem;
}

.small {
  font-size: 0.92rem;
}

.math.display {
  overflow-x: auto;
  padding: 0.25rem 0;
}

blockquote, .note-box {
  background-color: var(--light-blue);
  border-left: 5px solid var(--main-blue);
  padding: 0.8rem 1rem;
  border-radius: 8px;
}

hr {
  border: none;
  border-top: 1px solid var(--border-gray);
  margin: 1.5rem 0;
}


/* Presentation-style code chunks */
.chunk-id {
  display: inline-block;
  margin: 0.45rem 0 0.15rem 0;
  padding: 0.18rem 0.55rem;
  font-size: 0.78rem;
  font-weight: 700;
  color: #1f4e79;
  background: #e8f1f8;
  border: 1px solid #c6d7e6;
  border-radius: 999px;
  letter-spacing: 0.2px;
}

.cell:has(.sourceCode) {
  margin-top: 0.1rem;
  margin-bottom: 0.9rem;
}

div.sourceCode {
  display: inline-block;
  width: fit-content;
  max-width: 100%;
  margin-top: 0.1rem;
  margin-bottom: 0.4rem;
}

pre.sourceCode {
  margin: 0;
  padding: 0.55rem 0.75rem;
  border: 1px solid #cfd8e3;
  border-radius: 10px;
  background: #fbfcfe;
  box-shadow: 0 2px 8px rgba(31, 78, 121, 0.08);
  line-height: 1.25;
  font-size: 0.88rem;
  overflow-x: auto;
  white-space: pre-wrap;
}

pre.sourceCode code {
  white-space: pre-wrap;
  word-break: normal;
}

.cell-output, .cell-output-display {
  border-radius: 10px;
}

.result-table {
  font-size: 0.95rem;
}

</style>
# NYC Taxi Trip Total Payment Modellling
```{r setup}
#| echo: false
#| warning: false
#| message: false

rm(list = ls())

library(arrow)
library(readr)
library(dplyr)
library(ggplot2)
library(DT)
library(lubridate)
library(broom)
library(patchwork)
library(corrplot)

# Global plot style for the presentation
theme_set(
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 12),
      axis.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
)
```

# Objective:

- How are location, trip characteristics, payment type, trip duration, and fare-related charges associated with the total payment amount for NYC taxi trips?


# Data

```{r load-data}
#| cache: true
#| echo: false
#| warning: false
#| message: false

taxi_file <- "taxi_sample_20_percent.rds"
zones_file <- "taxi_zones.rds"

taxi_url <- "https://d37ci6vzurychx.cloudfront.net/trip-data/yellow_tripdata_2024-01.parquet"
zone_url <- "https://d37ci6vzurychx.cloudfront.net/misc/taxi_zone_lookup.csv"

if (file.exists(taxi_file) & file.exists(zones_file)) {
  
  taxi <- readRDS(taxi_file)
  zones <- readRDS(zones_file)
  
} else {
  
  taxi <- read_parquet(taxi_url)
  zones <- read_csv(zone_url, show_col_types = FALSE)
  
  set.seed(123)
  
  taxi <- taxi %>%
    slice_sample(prop = 0.20)
  
  saveRDS(taxi, taxi_file)
  saveRDS(zones, zones_file)
}
```

- **`Taxi`**:  January 2024 NYC Yellow Taxi Trip Dataset.

$$2,964,624 \times 19 \longrightarrow \text{ random subsample: }  296,462 \times 19$$




```{r}
#| echo: false
#| warning: false
#| message: false

taxi_ordered <- taxi %>%
  select(
    # Trip/provider information
    VendorID,
    store_and_fwd_flag,
    
    # Time information
    tpep_pickup_datetime,
    tpep_dropoff_datetime,
    
    # Location information
    PULocationID,
    DOLocationID,
    
    # Trip characteristics
    passenger_count,
    trip_distance,
    RatecodeID,
    
    # Payment information
    payment_type,
    
    # Fare and charges
    fare_amount,
    extra,
    mta_tax,
    tip_amount,
    tolls_amount,
    improvement_surcharge,
    total_amount,
    congestion_surcharge,
    Airport_fee
  )
```


```{r}
#| echo: false
#| warning: false
#| message: false

datatable(
  head(taxi_ordered, 4),
  caption = "First 4 rows of the taxi data, with variables grouped by category",
  options = list(
    scrollX = TRUE,
    pageLength = 4,
    dom = "t"
  ),
  rownames = FALSE,
  class = "stripe hover compact cell-border"
)
```




::: {.small}

| Category | Variables |
|---|---|
| **Trip and provider information** | **`VendorID`**, **`store_and_fwd_flag`** |
| **Time information** | **`tpep_pickup_datetime`**, **`tpep_dropoff_datetime`** |
| **Location information** | **`PULocationID`**, **`DOLocationID`** |
| **Trip characteristics** | **`passenger_count`**, **`trip_distance`**, **`RatecodeID`** |
| **Payment information** | **`payment_type`** |
| **Fare and related charges** | **`fare_amount`**, **`extra`**, **`mta_tax`**, **`tip_amount`**, **`tolls_amount`**, **`improvement_surcharge`**, **`total_amount`**, **`congestion_surcharge`**, **`Airport_fee`** |

:::





- **`zones`**: $\quad 265 \times 4$

```{r}
#| echo: false
#| warning: false
#| message: false

datatable(
  head(zones),
  caption = "First 4 rows of the zones data",
  options = list(
    scrollX = TRUE,
    pageLength = 4,
    dom = "t"
  ),
  rownames = FALSE,
  class = "stripe hover compact cell-border"
)
```


# Research Questions

**How are location, trip characteristics, payment type, trip duration, and non-fare charges associated with the total payment amount for NYC taxi trips?**

Specific research questions are:

1. Which variables are most strongly related to the total payment amount for a taxi trip?

2. How do trip distance and trip duration affect the total payment amount?

3. Do pickup and drop-off locations help explain differences in total payment amount?

4. Are airport trips, rush-hour trips, or different payment types associated with higher total payment amounts?

# Data Wrangling and Preprocessing

- Create the trip duration variable: **`trip_duration_min`**

$$trip\_duration\_min \textbf{ = } tpep\_dropoff\_datetime \quad \textbf{ - } \quad tpep\_pickup\_datetime$$

<div class="chunk-id">C01</div>

```{r}
#| cache: true
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi %>%
  mutate(
    trip_duration_min = as.numeric(
      difftime(
        tpep_dropoff_datetime,
        tpep_pickup_datetime,
        units = "mins"
      )
    )
  )
```


- Remove impossible or invalid trips:

<div class="chunk-id">C02</div>

```{r}
#| cache: true
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  filter(
    trip_duration_min > 0,
    trip_distance > 0,
    fare_amount > 0,
    passenger_count >= 1,
    passenger_count <= 6
  )
```

- Apply reasonable limits

<div class="chunk-id">C03</div>

```{r}
#| cache: true
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  filter(
    trip_duration_min <= 180,
    trip_distance <= 100,
    fare_amount <= 500
  )
```


- Avoid data leakage

Since **`total_amount`** is the response variable, including **`fare_amount`** as a predictor may create **data leakage**, because **`fare_amount`** is a direct component of the total payment calculation. Therefore, I remove **`fare_amount`** from the regression predictors.

<div class="chunk-id">C04</div>

```{r}
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  select(
    total_amount,
    trip_distance,
    tpep_pickup_datetime,
    trip_duration_min,
    passenger_count,
    tip_amount,
    tolls_amount,
    payment_type,
    PULocationID,
    DOLocationID
  )
```



- Check skewness and possible outliers for the main numeric variables:

```{r}
#| echo: false
#| warning: false
#| message: false

main_vars <- c(
  "trip_duration_min",
  "trip_distance",
  "tip_amount",
  "tolls_amount",
  "total_amount"
)

taxi_plot <- taxi_clean %>%
  select(all_of(main_vars)) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )
```
```{r}
#| cache: true
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 7

taxi_plot_zoom <- taxi_plot %>%
  group_by(variable) %>%
  filter(
    value >= quantile(value, 0.01, na.rm = TRUE),
    value <= quantile(value, 0.99, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(taxi_plot_zoom, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 50,
    fill = "#2C7FB8",
    color = "white"
  ) +
  geom_density(
    color = "#D62728",
    linewidth = 1.3
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "Histograms of Main Numeric Variables",
    subtitle = "Displayed after removing the bottom 1% and top 1% only for visualization",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 17, color = "gray30"),
    axis.title = element_text(face = "bold", size = 18),
    axis.text = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 18),
    strip.background = element_rect(fill = "#EEF5FB", color = "#D9E2EC"),
    panel.grid.minor = element_blank()
  )
```

```{r}
n_before <- nrow(taxi_clean)

taxi_clean <- taxi_clean %>%
  filter(
    tip_amount <= 8,
    tolls_amount < 4,
    total_amount <= 65,
    trip_distance <= 8,
    trip_duration_min <= 52
  )

n_after <- nrow(taxi_clean)

cat("Rows before filtering:", n_before, "\n")
cat("Rows after filtering:", n_after, "\n")
cat("Rows removed:", n_before - n_after, "\n")
```

```{r}
library(dplyr)
library(tidyr)

taxi_plot <- taxi_clean %>%
  select(
    tip_amount,
    tolls_amount,
    total_amount,
    trip_distance,
    trip_duration_min
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )
```

```{r histograms-after-outlier-filtering}
#| cache: true
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 7

library(dplyr)
library(tidyr)
library(ggplot2)

taxi_plot <- taxi_clean %>%
  select(
    tip_amount,
    tolls_amount,
    total_amount,
    trip_distance,
    trip_duration_min
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

ggplot(taxi_plot, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 50,
    fill = "#2C7FB8",
    color = "white"
  ) +
  geom_density(
    color = "#D62728",
    linewidth = 1.3
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "Histograms of Main Numeric Variables After Outlier Filtering",
    subtitle = "Based on the filtered dataset used for modeling",
    x = "Value",
    y = "Density"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    plot.subtitle = element_text(size = 17, color = "gray30"),
    axis.title = element_text(face = "bold", size = 18),
    axis.text = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 18),
    strip.background = element_rect(fill = "#EEF5FB", color = "#D9E2EC"),
    panel.grid.minor = element_blank()
  )
```



- Example: remove outliers for **`trip_duration_min`** using the IQR method:

The IQR rule keeps observations within the following range:

$$
LF \leq \text{observation} \leq UF
$$

where

$$
\begin{aligned}
LF &= Q_1 - 1.5 \times IQR, \\
UF &= Q_3 + 1.5 \times IQR.
\end{aligned}
$$

```{r}
#| echo: false
#| warning: false
#| message: false


Q1 <- quantile(taxi_clean$trip_duration_min, 0.25, na.rm = TRUE)
Q3 <- quantile(taxi_clean$trip_duration_min, 0.75, na.rm = TRUE)

IQR_value <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

taxi_clean <- taxi_clean %>%
  filter(
    trip_duration_min >= lower_bound,
    trip_duration_min <= upper_bound
  )
```

- Apply the same IQR method to the other main numeric variables:



```{r}
#| echo: false
#| warning: false
#| message: false

remove_iqr_outliers <- function(data, variable) {
  
  Q1 <- quantile(data[[variable]], 0.25, na.rm = TRUE)
  Q3 <- quantile(data[[variable]], 0.75, na.rm = TRUE)
  
  IQR_value <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  
  data %>%
    filter(
      .data[[variable]] >= lower_bound,
      .data[[variable]] <= upper_bound
    )
}

taxi_clean <- taxi_clean %>%
  remove_iqr_outliers("trip_distance") %>%
  remove_iqr_outliers("tip_amount") %>%
  remove_iqr_outliers("total_amount")
```



```{r}
#| echo: false
#| warning: false
#| message: false

main_vars <- c(
  "trip_duration_min",
  "trip_distance",
  "tip_amount",
  "tolls_amount",
  "total_amount"
)

taxi_plot <- taxi_clean %>%
  select(all_of(main_vars)) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )
```


- Apply a log transformation to the main skewed variables except **`tip_amount`** and **`tolls_amount`**:


<div class="chunk-id">C05</div>

```{r}
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  mutate(
    log_trip_duration_min = log1p(trip_duration_min),
    log_trip_distance = log1p(trip_distance),
    log_total_amount = log1p(total_amount)
  )
```

- Apply the same IQR outlier rule again after the log transformation. The rest of the analysis uses this updated filtered dataset.

```{r log-iqr-outlier-filtering}
#| cache: true
#| echo: false
#| warning: false
#| message: false

log_vars_for_iqr <- c(
  "log_trip_duration_min",
  "log_trip_distance",
  "log_total_amount"
)

remove_iqr_outliers_with_bounds <- function(data, variable) {
  
  Q1 <- quantile(data[[variable]], 0.25, na.rm = TRUE)
  Q3 <- quantile(data[[variable]], 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  
  lower_bound <- Q1 - 2.5 * IQR_value
  upper_bound <- Q3 + 2.5 * IQR_value
  
  data %>%
    filter(
      .data[[variable]] >= lower_bound,
      .data[[variable]] <= upper_bound
    )
}

n_before_log_iqr <- nrow(taxi_clean)

for (v in log_vars_for_iqr) {
  taxi_clean <- remove_iqr_outliers_with_bounds(taxi_clean, v)
}

n_after_log_iqr <- nrow(taxi_clean)

log_iqr_filter_summary <- data.frame(
  Step = "IQR filtering after log transformation",
  Rows_Before = n_before_log_iqr,
  Rows_After = n_after_log_iqr,
  Rows_Removed = n_before_log_iqr - n_after_log_iqr
)

DT::datatable(
  log_iqr_filter_summary,
  caption = "Rows Removed by IQR Filtering After Log Transformation",
  rownames = FALSE,
  options = list(dom = "t")
)
```

- Check histograms after log transformation:

```{r}
#| echo: false
#| warning: false
#| message: false

log_vars <- c(
  "log_trip_duration_min",
  "log_trip_distance",
  "log_total_amount"
)

taxi_plot_log <- taxi_clean %>%
  select(all_of(log_vars)) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6.5

taxi_clean <- taxi_clean %>%
  mutate(
    log_trip_duration_min = log1p(trip_duration_min),
    log_trip_distance = log1p(trip_distance),
    log_total_amount = log1p(total_amount)
  )

log_vars <- c(
  "log_trip_duration_min",
  "log_trip_distance",
  "log_total_amount"
)

taxi_plot_log <- taxi_clean %>%
  select(all_of(log_vars)) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "value"
  )

ggplot(taxi_plot_log, aes(x = value)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 50,
    fill = "#2C7FB8",
    color = "white"
  ) +
  geom_density(
    color = "#D62728",
    linewidth = 1.3
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "Histograms of Log-Transformed Variables",
    x = "Log(1 + Value)",
    y = "Density"
  ) +
  theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", size = 24),
    axis.title = element_text(face = "bold", size = 18),
    axis.text = element_text(size = 15),
    strip.text = element_text(face = "bold", size = 18),
    strip.background = element_rect(fill = "#EEF5FB", color = "#D9E2EC"),
    panel.grid.minor = element_blank()
  )
```


- Handle missing values

<div class="chunk-id">C06</div>

```{r}
#| echo: true
#| warning: false
#| message: false

missing_check <- data.frame(
  Check = "At least one missing value",
  Result = any(is.na(taxi_clean)),
  Total_Missing_Values = sum(is.na(taxi_clean))
)

DT::datatable(
  missing_check,
  rownames = FALSE
)
```
- Remove duplicated rows

<div class="chunk-id">C07</div>

```{r}
#| echo: true
#| warning: false
#| message: false

duplicate_check <- data.frame(
  Check = "Number of duplicated rows",
  Result = sum(duplicated(taxi_clean))
)

DT::datatable(
  duplicate_check,
  rownames = FALSE,
)
```
<div class="chunk-id">C08</div>

```{r}
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  distinct()
```

- Create time-based variables:

<div class="chunk-id">C09</div>

```{r}
#| echo: true
#| warning: false
#| message: false

library(lubridate)

taxi_clean <- taxi_clean %>%
  mutate(
    pickup_hour = hour(tpep_pickup_datetime),
    pickup_day = wday(tpep_pickup_datetime, label = TRUE, week_start = 1),
    weekend = ifelse(pickup_day %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(pickup_hour %in% c(7, 8, 9, 16, 17, 18), 1, 0),
    nighttime = ifelse(pickup_hour >= 22 | pickup_hour <= 5, 1, 0)
  )
```

- Merge the taxi zone lookup table

<div class="chunk-id">C10</div>

```{r}
#| cache: true
#| echo: true
#| warning: false
#| message: false

pickup_zones <- zones %>%
  rename(
    PULocationID = LocationID,
    pickup_borough = Borough,
    pickup_zone = Zone
  )

dropoff_zones <- zones %>%
  rename(
    DOLocationID = LocationID,
    dropoff_borough = Borough,
    dropoff_zone = Zone
  )

taxi_clean <- taxi_clean %>%
  left_join(pickup_zones, by = "PULocationID") %>%
  left_join(dropoff_zones, by = "DOLocationID")
```

- Remove unknown or missing boroughs

<div class="chunk-id">C11</div>

```{r}
#| echo: true
#| warning: false
#| message: false

pickup_borough_counts <- taxi_clean %>%
  count(pickup_borough, name = "count")

DT::datatable(
  pickup_borough_counts,
  caption = "Pickup Borough Counts",
  rownames = FALSE,
)

```

<div class="chunk-id">C12</div>

```{r}
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  filter(
    pickup_borough != "Unknown",
    dropoff_borough != "Unknown",
    !is.na(pickup_borough),
    !is.na(dropoff_borough)
  )
```

- Convert categorical variables to factors

<div class="chunk-id">C13</div>

```{r}
#| echo: true
#| warning: false
#| message: false

taxi_clean <- taxi_clean %>%
  mutate(
    pickup_day = as.factor(pickup_day),
    pickup_borough = as.factor(pickup_borough),
    dropoff_borough = as.factor(dropoff_borough),
    payment_type = as.factor(payment_type)
  )
```


# EDA

```{r}
#| cache: true
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 6

regression_data <- taxi_clean %>% 
  select(
    log_total_amount,
    log_trip_distance,
    log_trip_duration_min,
    passenger_count,
    tip_amount,
    tolls_amount,
    payment_type,
    pickup_borough,
    dropoff_borough,
    weekend,
    rush_hour,
    nighttime
  ) %>% 
  na.omit()

num_vars <- regression_data %>% 
  select(
    log_total_amount,
    log_trip_distance,
    log_trip_duration_min,
    passenger_count,
    tip_amount,
    tolls_amount
  )

cor_matrix <- cor(num_vars, use = "complete.obs")

corrplot::corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  tl.col = "black",
  tl.cex = 0.85,
  number.cex = 0.75,
  mar = c(0, 0, 2, 0),
  title = "Correlation Matrix of Key Numeric Variables"
)
```

```{r}
#| cache: true
#| echo: false
#| warning: false
#| message: false

set.seed(123)

sample_size <- min(10000, nrow(regression_data))

plot_data <- regression_data %>%
  slice_sample(n = sample_size)
```


```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6.5

presentation_theme <- theme_minimal(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", size = 22),
    plot.subtitle = element_text(size = 17, color = "gray30"),
    axis.title = element_text(face = "bold", size = 18),
    axis.text = element_text(size = 15),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(size = 14, color = "gray35", hjust = 0),
    plot.margin = margin(10, 15, 10, 15)
  )

p1 <- ggplot(
  plot_data,
  aes(x = log_trip_distance, y = log_total_amount)
) +
  geom_point(alpha = 0.15, size = 0.9, color = "#2C7FB8") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "#D62728") +
  labs(
    title = "Trip Distance and Total Payment",
    subtitle = "Longer trips are associated with higher total payment amounts",
    x = "Log(1 + Trip Distance)",
    y = "Log(1 + Total Amount)",
    caption = "Based on a random sample of 10,000 trips."
  ) +
  presentation_theme

p2 <- ggplot(
  plot_data,
  aes(x = log_trip_duration_min, y = log_total_amount)
) +
  geom_point(alpha = 0.15, size = 0.9, color = "#F28E2B") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, color = "#D62728") +
  labs(
    title = "Trip Duration and Total Payment",
    subtitle = "Longer trip durations are associated with higher total payment amounts",
    x = "Log(1 + Trip Duration in Minutes)",
    y = "Log(1 + Total Amount)",
    caption = "Based on a random sample of 10,000 trips."
  ) +
  presentation_theme

p1 + p2
```


```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 12
#| fig-height: 6.5

p3 <- ggplot(
  plot_data,
  aes(x = pickup_borough, y = log_total_amount)
) +
  geom_boxplot(
    aes(fill = pickup_borough),
    alpha = 0.75,
    width = 0.65,
    outlier.alpha = 0.15,
    outlier.size = 0.8
  ) +
  labs(
    title = "Total Payment by Pickup Borough",
    subtitle = "Distribution of total payment differs across pickup locations",
    x = "Pickup Borough",
    y = "Log(1 + Total Amount)",
    caption = "Based on a random sample of 10,000 trips."
  ) +
  presentation_theme +
  theme(
    axis.text.x = element_text(size = 15, angle = 35, hjust = 1)
  )

p4 <- ggplot(
  plot_data,
  aes(x = factor(rush_hour), y = log_total_amount)
) +
  geom_boxplot(
    aes(fill = factor(rush_hour)),
    alpha = 0.75,
    width = 0.55,
    outlier.alpha = 0.15,
    outlier.size = 0.8
  ) +
  scale_x_discrete(
    labels = c("0" = "Non-Rush Hour", "1" = "Rush Hour")
  ) +
  labs(
    title = "Total Payment by Rush-Hour Status",
    subtitle = "Comparison of payment distributions by time category",
    x = "Time Category",
    y = "Log(1 + Total Amount)",
    caption = "Based on a random sample of 10,000 trips."
  ) +
  presentation_theme

p3 + p4
```



# Modeling Objective

- The goal is to predict **`log_total_amount`**, defined as

$$
\log(1 + \text{total\_amount})
$$

using trip characteristics, payment type, location variables, and time indicators.

- A clean modeling data set and one common train-test split are used for all models.

- OLS is fitted first as the baseline model and evaluated using RMSE, test-set $R^2$, residual diagnostics, the Breusch-Pagan test, and VIF/GVIF.

- Because OLS shows nonconstant variance, robust standard errors and a high-leverage sensitivity check are considered.

- Ridge, LASSO, PCR, WLS regression, and $C_p$-selected OLS are compared using the same test-set RMSE and test-set $R^2$.

- The final model is selected based on prediction accuracy, simplicity, and interpretability.

## Setup for Modeling

```{r modeling-setup}
#| warning: false
#| message: false
#| echo: false

library(dplyr)
library(ggplot2)
library(DT)
library(broom)
library(car)
library(lmtest)
library(sandwich)
library(glmnet)
library(pls)
library(quantreg)

compute_metrics <- function(observed, predicted) {
  residual <- observed - predicted
  rmse <- sqrt(mean(residual^2))
  test_r2 <- 1 - sum(residual^2) / sum((observed - mean(observed))^2)
  
  data.frame(
    RMSE = round(rmse, 4),
    Test_R2 = round(test_r2, 4)
  )
}
```

## Data for Modeling

```{r create-modeling-data}
#| cache: true
#| echo: false
#| warning: false
#| message: false

regression_data_2 <- taxi_clean %>%
  select(
    log_total_amount,
    log_trip_distance,
    log_trip_duration_min,
    passenger_count,
    tip_amount,
    tolls_amount,
    payment_type,
    pickup_borough,
    dropoff_borough,
    weekend,
    rush_hour,
    nighttime
  ) %>%
  na.omit()
```

```{r}
#| warning: false
#| message: false
regression_data_3 <- regression_data_2 %>%
  filter(
    pickup_borough != "N/A",
    dropoff_borough != "N/A"
  ) %>%
  droplevels()
```



```{r train-test-split}
#| cache: true
#| echo: false
#| warning: false
#| message: false

set.seed(123)

n <- nrow(regression_data_3)

train_index <- sample(
  seq_len(n),
  size = 0.80 * n
)

train_data <- regression_data_3[train_index, ]
test_data  <- regression_data_3[-train_index, ]

split_size_table <- data.frame(
  Dataset = c("Training data (80%)", "Test data (20%)"),
  Number_of_Rows = c(nrow(train_data), nrow(test_data))
)

DT::datatable(
  split_size_table,
  caption = "Training and Test Set Sizes",
  rownames = FALSE,
  options = list(dom = "t")
)
```

# OLS Model

$$
\begin{aligned}
\log(1+\text{total\_amount})
= \beta_0 &+ \beta_1 \log(1+\text{trip\_distance}) \\[0.4em]
&\quad + \beta_2 \log(1+\text{trip\_duration\_min}) \\[0.4em]
&\quad\quad + \beta_3 \text{passenger\_count} \\[0.4em]
&\quad\quad\quad + \beta_4 \text{tip\_amount} \\[0.4em]
&\quad\quad\quad\quad + \beta_5 \text{tolls\_amount} \\[0.4em]
&\quad\quad\quad\quad\quad + \beta_6 \text{weekend} \\[0.4em]
&\quad\quad\quad\quad\quad\quad + \beta_7 \text{rush\_hour} \\[0.4em]
&\quad\quad\quad\quad\quad\quad\quad + \beta_8 \text{nighttime} \\[0.4em]
&\quad\quad\quad\quad\quad\quad\quad\quad + \boldsymbol{\gamma}^\top \text{payment\_type} \\[0.4em]
&\quad\quad\quad\quad\quad\quad\quad\quad\quad + \boldsymbol{\delta}^\top \text{pickup\_borough} \\[0.4em]
&\quad\quad\quad\quad\quad\quad\quad\quad\quad\quad + \boldsymbol{\eta}^\top \text{dropoff\_borough} + \varepsilon .
\end{aligned} 
$$

```{r ols-fit}
#| cache: true
#| echo: false
#| warning: false
#| message: false

ols_model <- lm(
  log_total_amount ~ log_trip_distance +
    log_trip_duration_min +
    passenger_count +
    tip_amount +
    tolls_amount +
    payment_type +
    pickup_borough +
    dropoff_borough +
    weekend +
    rush_hour +
    nighttime,
  data = train_data
)

ols_coef_table <- tidy(ols_model) %>%
  mutate(across(where(is.numeric), ~ round(.x, 4)))

DT::datatable(
  ols_coef_table,
  caption = "OLS Regression Coefficients",
  rownames = FALSE,
  options = list(scrollX = TRUE, pageLength = 10)
)
```

## OLS Test-Set Evaluation

```{r}
#| cache: true
#| echo: false
#| warning: false
#| message: false
#| 
# Categorical variables used in the model
cat_vars <- c("payment_type", "pickup_borough", "dropoff_borough")

# Remove test rows with factor levels not seen in training
for (v in cat_vars) {
  
  train_levels <- unique(train_data[[v]])
  
  test_data <- test_data %>%
    filter(.data[[v]] %in% train_levels)
  
  train_data[[v]] <- factor(train_data[[v]], levels = train_levels)
  test_data[[v]]  <- factor(test_data[[v]], levels = train_levels)
}
```


```{r ols-evaluation}
#| cache: true
#| echo: false
#| warning: false
#| message: false

ols_pred <- predict(
  ols_model,
  newdata = test_data
)

ols_results <- data.frame(
  observed = test_data$log_total_amount,
  predicted = ols_pred,
  residual = test_data$log_total_amount - ols_pred
)

ols_metrics <- compute_metrics(
  observed = ols_results$observed,
  predicted = ols_results$predicted
)

ols_rmse <- ols_metrics$RMSE
ols_test_r2 <- ols_metrics$Test_R2

ols_eval_table <- data.frame(
  Model = "OLS",
  ols_metrics
)

DT::datatable(
  ols_eval_table,
  caption = "OLS Test-Set Evaluation Metrics",
  rownames = FALSE,
  options = list(dom = "t")
)
```

```{r ols-predicted-observed-plot}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 6

ggplot(ols_results, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.20, size = 0.8) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  labs(
    title = "OLS: Predicted vs Observed Values",
    x = "Observed Log(1 + Total Amount)",
    y = "Predicted Log(1 + Total Amount)"
  ) +
  theme_minimal(base_size = 16)
```


## OLS Diagnostic Checks


- Residuals vs Fitted Values

```{r ols-residual-plot}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 6

ols_diag <- data.frame(
  fitted = fitted(ols_model),
  residual = resid(ols_model)
)

ggplot(ols_diag, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_hline(
    yintercept = 0,
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  labs(
    title = "OLS: Residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Residuals"
  ) +
  theme_minimal(base_size = 16)
```

- Test for Constant Variance

```{r ols-bp-test}
#| warning: false
#| message: false
#| echo: false

bp_test <- bptest(ols_model)

bp_table <- data.frame(
  Test = "Breusch-Pagan test",
  Statistic = round(as.numeric(bp_test$statistic), 4),
  df = as.numeric(bp_test$parameter),
  p_value = signif(bp_test$p.value, 4)
)

DT::datatable(
  bp_table,
  caption = "Breusch-Pagan Test for Constant Variance",
  rownames = FALSE,
  options = list(dom = "t")
)
```


- Multicollinearity Check Using VIF/GVIF

The model includes categorical: The adjusted value, $GVIF^{1/(2Df)}$.
output: true

```{r ols-gvif-adjusted-table}
#| warning: false
#| message: false
#| echo: false

library(car)
library(dplyr)
library(DT)

# Original OLS formula
ols_formula <- log_total_amount ~ log_trip_distance +
  log_trip_duration_min +
  passenger_count +
  tip_amount +
  tolls_amount +
  payment_type +
  pickup_borough +
  dropoff_borough +
  weekend +
  rush_hour +
  nighttime

# Make sure categorical variables are factors and unused levels are dropped
cat_vars <- c("payment_type", "pickup_borough", "dropoff_borough")

train_data <- train_data %>%
  mutate(
    across(all_of(cat_vars), as.factor),
    across(all_of(cat_vars), droplevels)
  )

# Fit initial OLS model
ols_model <- lm(
  ols_formula,
  data = train_data
)

# Identify aliased coefficients
aliased_coef <- names(coef(ols_model))[is.na(coef(ols_model))]

# If there are aliased coefficients, remove the corresponding model terms
if (length(aliased_coef) > 0) {
  
  x_mat <- model.matrix(ols_model)
  assign_vec <- attr(x_mat, "assign")
  term_labels <- attr(terms(ols_model), "term.labels")
  
  coef_to_term <- data.frame(
    Coefficient = colnames(x_mat),
    Term_Index = assign_vec
  ) %>%
    filter(Coefficient %in% aliased_coef, Term_Index > 0) %>%
    mutate(Term = term_labels[Term_Index])
  
  terms_to_drop <- unique(coef_to_term$Term)
  
  remaining_terms <- setdiff(term_labels, terms_to_drop)
  
  gvif_formula <- as.formula(
    paste("log_total_amount ~", paste(remaining_terms, collapse = " + "))
  )
  
  gvif_model <- lm(
    gvif_formula,
    data = train_data
  )
  
} else {
  
  terms_to_drop <- character(0)
  gvif_model <- ols_model
}

# Compute GVIF / Adjusted GVIF
vif_raw <- car::vif(gvif_model)

if (is.matrix(vif_raw)) {
  
  vif_table <- data.frame(
    Variable = rownames(vif_raw),
    GVIF = vif_raw[, "GVIF"],
    Df = vif_raw[, "Df"],
    Adjusted_GVIF = vif_raw[, "GVIF^(1/(2*Df))"]
  ) %>%
    mutate(
      GVIF = round(GVIF, 4),
      Df = round(Df, 0),
      Adjusted_GVIF = round(Adjusted_GVIF, 4)
    )
  
} else {
  
  vif_table <- data.frame(
    Variable = names(vif_raw),
    GVIF = as.numeric(vif_raw),
    Df = 1,
    Adjusted_GVIF = sqrt(as.numeric(vif_raw))
  ) %>%
    mutate(
      GVIF = round(GVIF, 4),
      Adjusted_GVIF = round(Adjusted_GVIF, 4)
    )
}

# Optional: show dropped aliased terms, if any
if (length(terms_to_drop) > 0) {
  dropped_terms_table <- data.frame(
    Dropped_Aliased_Term = terms_to_drop
  )
  
  DT::datatable(
    dropped_terms_table,
    caption = "Terms Removed Before Computing GVIF Due to Aliased Coefficients",
    rownames = FALSE,
    options = list(dom = "t")
  )
}

# Final GVIF table
DT::datatable(
  vif_table,
  caption = "OLS Multicollinearity Check Using GVIF and Adjusted GVIF",
  rownames = FALSE,
  options = list(scrollX = TRUE, pageLength = 15)
)
```


## Stablize residuals' variance

- Weighted Least Squares (WLS): failed
- Variable Selection (Mallows’ $C_p$): failed


```{r aligned-model-matrices}
#| cache: true
#| warning: false
#| message: false
#| echo: false
#| output: false

## Aligned Model Matrices for Regularized and Matrix-Based Models

model_formula <- log_total_amount ~ log_trip_distance +
  log_trip_duration_min +
  passenger_count +
  tip_amount +
  tolls_amount +
  payment_type +
  pickup_borough +
  dropoff_borough +
  weekend +
  rush_hour +
  nighttime

x_all <- model.matrix(
  model_formula,
  data = regression_data_3
)[, -1, drop = FALSE]

y_all <- regression_data_3$log_total_amount

x_train <- x_all[train_index, , drop = FALSE]
x_test  <- x_all[-train_index, , drop = FALSE]

y_train <- y_all[train_index]
y_test  <- y_all[-train_index]

matrix_dimension_table <- data.frame(
  Matrix = c("x_train", "x_test"),
  Rows = c(nrow(x_train), nrow(x_test)),
  Columns = c(ncol(x_train), ncol(x_test))
)

DT::datatable(
  matrix_dimension_table,
  caption = "Aligned Model Matrix Dimensions",
  rownames = FALSE,
  options = list(dom = "t")
)
```

## Ridge Regression

$$\hat{\boldsymbol{\beta}}_{\text{ridge}}
=
\arg\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \mathbf{x}_i^\top \boldsymbol{\beta}\right)^2
+
\lambda \sum_{j=1}^{p} \beta_j^2
\right\},
\qquad \lambda \ge 0.$$

```{r ridge-fit}
#| cache: true
#| warning: false
#| message: false
#| echo: false

set.seed(123)

ridge_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 0,
  nfolds = 10,
  standardize = TRUE
)

ridge_lambda_min <- ridge_cv$lambda.min
ridge_lambda_1se <- ridge_cv$lambda.1se

ridge_lambda_table <- data.frame(
  Lambda_Type = c("lambda.min", "lambda.1se"),
  Lambda_Value = c(ridge_lambda_min, ridge_lambda_1se)
)

DT::datatable(
  ridge_lambda_table,
  caption = "Ridge Cross-Validated Lambda Values",
  rownames = FALSE,
  options = list(dom = "t")
)
```

```{r ridge-cv-plot}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 6

plot(ridge_cv)
```

```{r ridge-coefficients}
#| warning: false
#| message: false
#| echo: false

ridge_coef <- coef(ridge_cv, s = ridge_lambda_min)

ridge_coef_table <- data.frame(
  Variable = rownames(ridge_coef),
  Coefficient = as.numeric(ridge_coef)
) %>%
  mutate(Coefficient = round(Coefficient, 4))

DT::datatable(
  ridge_coef_table,
  caption = "Ridge Regression Coefficients Using lambda.min",
  rownames = FALSE,
  options = list(scrollX = TRUE, pageLength = 15)
)
```

```{r ridge-evaluation}
#| cache: true
#| warning: false
#| message: false
#| echo: false

ridge_pred <- predict(
  ridge_cv,
  s = ridge_lambda_min,
  newx = x_test
)

ridge_metrics <- compute_metrics(
  observed = y_test,
  predicted = as.numeric(ridge_pred)
)

ridge_rmse <- ridge_metrics$RMSE
ridge_test_r2 <- ridge_metrics$Test_R2

ridge_eval_table <- data.frame(
  Model = "Ridge",
  ridge_metrics
)

DT::datatable(
  ridge_eval_table,
  caption = "Ridge Test-Set Evaluation Metrics",
  rownames = FALSE,
  options = list(dom = "t")
)
```

Ridge regression is evaluated mainly through prediction performance. Since the adjusted GVIF values did not show severe multicollinearity, Ridge is not expected to greatly improve over OLS.

## LASSO Regression

LASSO is another regularized model. Unlike Ridge, LASSO can shrink some coefficients exactly to zero, so it can perform variable selection.

$$
\hat{\boldsymbol{\beta}}_{\text{lasso}}
=
\arg\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(y_i - \mathbf{x}_i^\top \boldsymbol{\beta}\right)^2
+
\lambda \sum_{j=1}^{p} |\beta_j|
\right\},
\qquad \lambda \ge 0.
$$

```{r lasso-fit}
#| cache: true
#| warning: false
#| message: false
#| echo: false

set.seed(123)

lasso_cv <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,
  nfolds = 10,
  standardize = TRUE
)

lasso_lambda_min <- lasso_cv$lambda.min
lasso_lambda_1se <- lasso_cv$lambda.1se

lasso_lambda_table <- data.frame(
  Lambda_Type = c("lambda.min", "lambda.1se"),
  Lambda_Value = c(lasso_lambda_min, lasso_lambda_1se)
)

DT::datatable(
  lasso_lambda_table,
  caption = "LASSO Cross-Validated Lambda Values",
  rownames = FALSE,
  options = list(dom = "t")
)
```

```{r lasso-cv-plot}
#| warning: false
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 6

plot(lasso_cv)
```

```{r lasso-coefficients}
#| warning: false
#| message: false
#| echo: false

lasso_coef <- coef(lasso_cv, s = lasso_lambda_min)

lasso_coef_table <- data.frame(
  Variable = rownames(lasso_coef),
  Coefficient = as.numeric(lasso_coef)
) %>%
  mutate(Coefficient = round(Coefficient, 4))

DT::datatable(
  lasso_coef_table,
  caption = "LASSO Regression Coefficients Using lambda.min",
  rownames = FALSE,
  options = list(scrollX = TRUE, pageLength = 15)
)

lasso_nonzero_coef_table <- lasso_coef_table %>%
  filter(Coefficient != 0)

DT::datatable(
  lasso_nonzero_coef_table,
  caption = "Nonzero LASSO Coefficients Using lambda.min",
  rownames = FALSE,
  options = list(scrollX = TRUE, pageLength = 15)
)
```

```{r lasso-evaluation}
#| cache: true
#| warning: false
#| message: false
#| echo: false

lasso_pred <- predict(
  lasso_cv,
  s = lasso_lambda_min,
  newx = x_test
)

lasso_metrics <- compute_metrics(
  observed = y_test,
  predicted = as.numeric(lasso_pred)
)

lasso_rmse <- lasso_metrics$RMSE
lasso_test_r2 <- lasso_metrics$Test_R2

lasso_eval_table <- data.frame(
  Model = "LASSO",
  lasso_metrics
)

DT::datatable(
  lasso_eval_table,
  caption = "LASSO Test-Set Evaluation Metrics",
  rownames = FALSE,
  options = list(dom = "t")
)
```

LASSO is useful if it achieves similar prediction accuracy to OLS while keeping a simpler model with fewer active predictors.

# Principal Component Regression

### Principal Component Regression (PCR)

PCR replaces correlated predictors with uncorrelated principal components.

Each principal component is a weighted combination of the original predictors:

$$
PC_k = v_{k1}X_1 + v_{k2}X_2 + \cdots + v_{kp}X_p
$$

$$
\text{Correlated: $X=[X_1, X_2, \dots, X_p]$ $\xrightarrow{\text{PC Analysis}}$ Uncorrelated: $Z=[Z_1, Z_2, \dots, Z_p]$}
$$


The PCR model is then

$$
y = Z\beta_z + \epsilon.
$$

After centering and scaling $X$, PCA is performed using singular value decomposition:

$$
X = UDV'
$$

where $V$ contains the principal component directions. The principal component matrix is

$$
Z = XV.
$$

OLS on the original predictors is

$$
\hat{\beta}_x = (X'X)^{-1}X'y.
$$

OLS on the principal components is

$$
\hat{\beta}_z = (Z'Z)^{-1}Z'y.
$$

Using all principal components gives the same fitted values as OLS:

$$
\hat{y} = X\hat{\beta}_x = Z\hat{\beta}_z.
$$

Therefore, PCR removes multicollinearity by using uncorrelated components, while keeping the same predictions when all components are used.

```{r pcr-matrix-preparation}
#| cache: true
#| warning: false
#| message: false
#| echo: false
#| output: false

x_train_sds_check <- apply(x_train, 2, sd)
near_zero_var_cols <- x_train_sds_check < 1e-8

x_train_pcr <- x_train[, !near_zero_var_cols, drop = FALSE]
x_test_pcr  <- x_test[, !near_zero_var_cols, drop = FALSE]

x_train_means <- colMeans(x_train_pcr)
x_train_sds <- apply(x_train_pcr, 2, sd)

x_train_pcr_scaled <- scale(
  x_train_pcr,
  center = x_train_means,
  scale = x_train_sds
)

x_test_pcr_scaled <- scale(
  x_test_pcr,
  center = x_train_means,
  scale = x_train_sds
)

pcr_matrix_check <- data.frame(
  Matrix = c("Training PCR matrix", "Test PCR matrix"),
  Non_Finite_Values = c(
    sum(!is.finite(x_train_pcr_scaled)),
    sum(!is.finite(x_test_pcr_scaled))
  )
)

DT::datatable(
  pcr_matrix_check,
  caption = "PCR Matrix Check",
  rownames = FALSE,
  options = list(dom = "t")
)
```

```{r pcr-fit}
#| cache: true
#| warning: false
#| message: false
#| echo: true

set.seed(123)

pcr_model <- pcr(
                    y_train  ~  x_train_pcr_scaled,
  scale = FALSE,
  validation = "CV"
)
```

In the earlier output, the full 21-component result was numerically unstable. Therefore, 20 components are selected to avoid the unstable final component while retaining almost all predictor variation.

## Which PC's should be involved in the model?

- Elbow method

```{r pcr-validation-plot}
#| cache: true
#| warning: false
#| message: false
#| echo: false
#| fig-width: 8
#| fig-height: 6

# Define number of PCR components inside the chunk
pcr_ncomp <- min(17, pcr_model$ncomp)

validationplot(
  pcr_model,
  val.type = "MSEP",
  estimate = "CV",
  ncomp = 1:pcr_ncomp,  # do not include 0 here; validationplot adds the 0-component model automatically
  ylim = c(0, 0.1),
  xaxt = "n",
  type = "b",
  lty = 2,
  col = "blue",
  pch = 19,
  main = "Cross-Validated MSEP for PCR by Number of Components"
)

axis(
  side = 1,
  at = 0:pcr_ncomp,
  labels = 0:pcr_ncomp
)
```


```{r pcr-evaluation}
#| cache: true
#| warning: false
#| message: false
#| echo: false
pcr_ncomp <-3
pcr_pred <- predict(
  pcr_model,
  newdata = data.frame(x_train_pcr_scaled = I(x_test_pcr_scaled)),
  ncomp = pcr_ncomp
)

pcr_pred <- as.numeric(pcr_pred)

pcr_metrics <- compute_metrics(
  observed = y_test,
  predicted = pcr_pred
)

pcr_rmse <- pcr_metrics$RMSE
pcr_test_r2 <- pcr_metrics$Test_R2

pcr_eval_table <- data.frame(
  Model = "PCR",
  Components = pcr_ncomp,
  pcr_metrics
)

DT::datatable(
  pcr_eval_table,
  caption = "PCR Test-Set Evaluation Metrics",
  rownames = FALSE,
  options = list(dom = "t")
)
```



# Comparison and Conclusion

The final comparison uses the same test data and the same two metrics for all predictive models: RMSE and test-set $R^2$. Robust standard errors are not included as a separate predictive model because they change inference, not predictions.

```{r final-model-comparison}
#| cache: true
#| warning: false
#| message: false
#| echo: false

model_comparison <- data.frame(
  Model = c("OLS", "Ridge", "LASSO", "PCR"),
  RMSE = c(
    round(ols_rmse, 4),
    round(ridge_rmse, 4),
    round(lasso_rmse, 4),
    round(pcr_rmse, 4)
  ),
  Test_R2 = c(
    round(ols_test_r2, 4),
    round(ridge_test_r2, 4),
    round(lasso_test_r2, 4),
    round(pcr_test_r2, 4)
  )
)

DT::datatable(
  model_comparison,
  caption = "Final Model Comparison Table",
  rownames = FALSE,
  options = list(dom = "t")
)
```

- The initial objective was to address possible multicollinearity among predictors.

- After checking VIF/GVIF diagnostics, especially adjusted GVIF values, we found that multicollinearity was not a serious issue.

- Thus, Ridge, LASSO, and PCR were used mainly to compare prediction accuracy with OLS.

- Model diagnostics showed non-constant residual variance, so we shifted the focus toward heteroscedasticity.

- We tried several model improvements, including:
  - subset model selection,
  - PCA subset selection,
  - different outlier filters,
  - quadratic terms,
  - cubic terms,
  - interaction terms,
  - combinations of main, interaction, quadratic, and cubic terms.

- None of these approaches fully resolved the non-constant variance issue.

- Therefore, the final model is more reliable for prediction than interpretation.

- Coefficient interpretations and p-values should be treated with caution.

- The final focus is prediction accuracy and model comparison, not formal inference.

# Acknowledgement 

- ChatGPT
- Some of libraries:

```{r}
#| echo: true
#| warning: false
#| message: false

library(arrow)
library(readr)
library(dplyr)
library(ggplot2)
library(DT)
library(glmnet)
library(pls)
library(lubridate)
library(patchwork)   
library(corrplot)
library(tidyr)
```
